\Large Generation of Mathematical Programming Representation for Discrete Event Simulation Models of Timed Petri Nets

### 2022-10-30 Answer to reviewers.tex

\usepackage[natbibapa,nodoi]{apacite}
\setlength\bibhang{12pt}
\renewcommand\bibliographytypesize{\fontsize{10}{12}\selectfont}

\usepackage{color}

\begin{document}
\noindent
Dear DE, AE and reviewers,

Thanks to your valuable %review 
comments, this submission has undergone a major revision. In this %new 
revised version, the MP models are developed based on general timed Petri nets (TPN) instead of the proposed representation of discrete event dynamic systems (DEVS) that was used in the previous submission. This change reduces the difficulty of applying the proposed approach and can be better understood by people familiar with DEVS.

@@

The concerns about application, usefulness, benefit and computational complexity were mentioned by the editors and reviewers for several times.

To answer such concerns, we have seen many applications of the use of mathematical programming (MP) for the representation of system dynamics in literature. Most of the works couples the system dynamics and optimization, as in \cite{di2020makespan}, \cite{weiss2015buffer}, \cite{bemporad1999control}, \cite{alfieri2020time}. All those works solve problems in industrial applications, thus, the applicability has been proved. The properties of DEVS can also be analyzed through MPs, as in \cite{basile2012k}.

The usefulness %and benefit 
of this work can be justified in two aspects, namely generalization and convenience. TPN is a widely--used general modeling diagram of DEVS, and the proposed work would be easy to apply for people in this field.

The computational complexity of using MP to optimize DEVS has been %noticed 
addressed in many works, and they dealt the issue with classic MP techniques such as Benders decomposition, linear approximation, columns-row generation. We cannot deny that the computation issue really exists, but the ways to overcome are also vast. Furthermore, the translation is bidirectional. One can translate simulation into MP to optimize a system, and also translate the resulting MP into simulation. The value of translating MP into simulation is that the research on MP (such as duality and sensitivity) can %dig
increase the value of simulation, which is currently usually regarded only as a performance evaluation tool.

@@

In the following, we answer the comments in details.

\section{DE comments}
\textit{The merit of the proposed technique should be clarified either in a general setting or on specific classes of systems like Petri nets or finite automata, and \textcolor{blue}{comparisons with extant tools and algorithms} should be made.}

\noindent
\textbf{Answer:} %In this submission, 
In the revised manuscript our approach %works
has been reformulated to have it working under general settings deterministic timed Petri net.

@@

\section{AE Comments}
\textit{The contribution and usefulness of the proposed approach is not convincing and should be improved. }

\textit{1. The authors should convincingly demonstrate the benefit of the proposed MPR over other formalism like the automata and PN while also considering the complexity burden of solving the resulting mathematical program. }

\noindent
\textbf{Answer:} %In this submission, 
In the revised manuscript, we justify the benefit and computational burden of the proposed approach as follows:

The benefits of using MP model to represent discrete event systems are many. As already mentioned above, when coupled with optimization, the MP-based algorithms can be faster and %gain 
reach better solution than black--box optimization algorithms. Furthermore, black--box approaches have limited capability is solving constrained optimization problems, %but 
while MP-based approaches can easily deal with them (\cite{zhang2020models}). Third, the vast theoretical and methodological results developed in the MP field can be introduced into the study of DEVS through simulation. For instance, using sensitivity analysis of MP models, DES can %be
become a gradient estimation tool, as suggested in \cite{chan2008optimization}. In fact, we do not proposed to totally replace simulation with MP, but to enrich the toolbox for analyzing and optimizing DEVS. Finally, the application of MP models of DEVS is not limited to optimization. \cite{basile2012k} proposed the sufficient and necessary condition of K-diagnosability of TPN based on MP.

The major concern about the application of MP is the computational complexity, %and the solving procedure 
as the solution procedure can be time-consuming or even unbearable. However, many approaches from optimization community are available to improve the efficiency. Linear programming approximation (\cite{alfieri2012mathematical}), Benders decomposition (\cite{weiss2015buffer}), row-column generation (\cite{alfieri2020time}), have been studied to solve optimization problems in %reality
real contexts based-on MP model of DEVS.

@@

\noindent
\textbf{Answer:}

\section{Reviewer 1}
\textit{However, I think that the authors should clarify better their contribution with respect to the standard translation of other modelling formalisms into mathematical programming (see references below). From a certain point of view, mathematical programming is also a modelling formalism, even it is usually considered more a computational tool, anyway it is evident that it cannot replace finite state automata or Petri nets to model DES. Hence, my question is, there is a way to translate, as for example a PN using the proposed technique? If yes, is it more efficient? If not, what is the effective application domain of the proposed technique, in addition to queueing systems?}

\textit{References: \cite{basile2012k}, \cite{basile2012k}, \cite{bemporad1999control}}

\noindent
\textbf{Answer:} As the reviewer suggests, %this submission proposed 
in the revised manuscript we propose the translation of TPN into MP.

The difference between the proposed approach and the suggested references are %as follows.
detailed in the following. \cite{bemporad1999control} proposed a mixed integer programming modeling framework for hybrid systems, whose states are mixed integer, in discrete time, while this work deals with continuous--time discrete--state systems. An MP model was proposed to minimize the makespan of single-server manufacturing systems in \cite{di2020makespan}, %but this
while our work is more general and can model multi-server systems. \cite{basile2012k} proposed the model based on PN, not TPN, because %it deals with 
the paper is about diagnosability %other 
rather than time-dependent performance measures, such as waiting time or throughput.

The effective application domain of the proposed technique is basically related to optimization. The MP model of TPN %is easy to expand 
can be easily expanded to an MP model of TPN optimization. (\textcolor{blue}{to be expanded})

@@

\section{Reviewer 2}

\textit{1. The main criticisms I can to formulate about this MPR approach is that
(i) it produces only fixed-length simulations of a system, and (ii) it
results in high-dimension linear programs (O(NxE) where N is the
length of the execution and E the number of event types). This puts
stringent restrictions on the possible applications of this
encoding. For instance, steady-state behavior can only be
approximated, at the expense of very large MPR systems. This casts a
serious doubt about the usefulness of the whole approach, in
comparison with a DES simulator, implementing an operational
semantics of the particular DES considered in the paper.}

\noindent
\textbf{Answer:} The reviewer's comment is reasonable. We %justify 
discuss the computational burden as follows in the submission:

The major concern about the application of MP is the computational complexity, %and the solving procedure 
as the solution procedure can be time-consuming or even unbearable. However, many approaches from optimization community are available to improve the efficiency. Linear programming approximation (\cite{alfieri2012mathematical}), Benders decomposition (\cite{weiss2015buffer}), row-column generation (\cite{alfieri2020time}), have been studied to solve optimization problems in %reality
real contexts based-on MP model of DEVS.

### 2022-10-30 main.tex
Arianna revision

\title{\Large From black box to white box: automated generation of mathematical programming representation for discrete event simulation models}

@@

\bigskip
\begin{abstract}
This work proposes a mathematical programming (MP) representation of discrete event simulation of timed Petri nets (TPN). Currently, MP techniques are not widely applied to optimizing discrete event systems, and this work connects the two fruitful research fields. In the MP formalism, the decision variables are composed of the transition firing time and the markings, while the constraints represents the state transition logic and temporal sequences among events. The MP model and a simulation run of the TPN are totally equivalent, which is validated with several cases. Using a TPN model as input, the MP model can be routinely and conveniently generated.

@@

Discrete event dynamic systems (DEVS) has wide application in manufacturing and service applications. Even though many theoretical and analytical studies on discrete event dynamic systems have been developed, performance of real systems sometimes can only be evaluating with discrete event simulation (DES) due to system complexity. Hence, simulation--optimization algorithms are widely used when performance evaluation must be coupled with optimization, i.e., when the best system configuration, according to some criteria, has to be found meanwhile guaranteeing a given value of some performance measure.

Most of the simulation--optimization studies consider DES as a \textit{black--box} function. Under the black--box setting, simulation module and optimization module are decoupled, except that simulation module needs to provide the performance measures of certain solutions when the optimization module queries. With quite limited information from simulation, the optimization module must have many trials, so the black--box procedure can be computationally inefficient.

A promising direction to improve the inefficiency of black-box procedure is to merge the system dynamics into optimization, i.e., considering DES as /textit{white-box} and modeling system dynamics as part of mathematical programming (MP) models. Several works can be found in literature. An MP model was proposed to minimize the makespan of single-server manufacturing systems in \cite{di2020makespan}. The buffer allocation problem in multi-stage production flow lines with blocking is formulated as MP and solved in \cite{matta2008simulation}, \cite{weiss2015buffer} and \cite{alfieri2020time}. The optimization of systems with complex blocking mechanisms, for instance extended kanban, is addressed in \cite{pedrielli2015integrated} with MP techniques. Decomposition or linear approximation approaches are applied to solving the MP models, and it can be rather efficient than black-box algorithms and the optimality is also enhanced (\cite{weiss2015buffer}, \cite{alfieri2020time}).

However, white-box simulation optimization approaches are not as widely-applied as block-box ones. The reason is that developing a white-box algorithm requires one to master discrete event dynamic systems, its simulation and mathematical programming, but those disciplines are not tightly connected. The above-mentioned works are all tailored to specific cases, and there is a lack of generalization to extend the works to other problems. To fill this gap, this work proposes an approach to translating the simulation of general timed Petri net (TPN) into MP models.

In literature, a few works address the similar problem, i.e.,  translating the behavior of general discrete event dynamic systems into MP. \cite{bemporad1999control} proposed a mixed integer programming modeling framework for hybrid systems, whose states are mixed integer, in discrete time. Differently, this work deals with continuous--time discrete--state systems. \cite{chan2008optimization} proposed a modeling framework to translate a DES model into an MP model in a general sense based on the Event Relationship Graph (ERG) of the system dynamics. However, the ERG is not as commonly-used as TPN, and the approach is not used often.

The benefits of using MP model to present discrete event systems are many. As already mentioned above, when coupled with optimization, the MP-based algorithms can be faster and gain better solution than black--box optimization algorithms. Furthermore, black--box approaches have limited capability is solving constrained optimization problems, but MP-based approaches can easily deal with them \cite{zhang2020models}. Thirdly, the vast theoretical and methodological results developed in the MP field can be introduced into the study of DEVS through simulation. For instance, using sensitivity analysis of MP models, DES can be a gradient estimation tool, as suggested in \cite{chan2008optimization}. In fact, we do not proposed to totally replace simulation with MP, but to enrich the toolbox for analyzing and optimizing DEVS. Finally, the application of MP models of DEVS is not limited to optimization. \cite{basile2012k} proposed the sufficient and necessary condition of K-diagnosability of TPN based on MP.

The major concern about the application of MP is the computational complexity, and the solving procedure can be time-consuming or even unbearable. However, many approaches from optimization community are available to improve the efficiency. Linear programming approximation (\cite{alfieri2012mathematical}), Benders decomposition (\cite{weiss2015buffer}), row-column generation (\cite{alfieri2020time}), have been studied to solve optimization problems in reality based-on MP model of DEVS.

The contributions of this work is to propose the first framework translating general TPN into MP models. The translation procedure can be conveniently implemented in general--purpose programming languages. This work lays a foundation of developing MP-based approaches for analyzing and optimizing general TPN, based on the vast literature in MP field. 
%The proposed MP models can also be easily transformed into an MPR for optimizing the design parameters of the DES, which are common optimization problems in operations management field.

The rest of the paper is organized as follows. Section \ref{sec:sim} briefly introduce the TPN and its simulation algorithm. Section \ref{sec:TPN_MPR} describes the generation of the MP of a TPN model. Section XXX shows several examples of application and validation of the equivalence of simulation and the MP. Discussion and conclusion are reported in Section XXX and XXX, respectively.

@@

\section{Simulating TPN} \label{sec:sim}

\subsection{Introduction to timed Petri nets}
A TPN is a directed bipartite graph. The set of nodes in that graph is partitioned into a set of \textit{places} $P$ and a set of \textit{transitions} $T$. Places and Transitions are connected with weighted and directed edges. As in the right side of Figure \ref{fig:gg1_petri}, places (i.e., $p_1$, $p_2$, $p_3$), transitions (i.e., $t_1$, $t_2$) and tokens are represented by white circles, rectangles and small black circles, respectively. Places hold \textit{tokens}, and tokens are transferred through transition \textit{firings}, i.e., one transition firing absorbs tokens from some places and releases to some other places. Each transition $t\in{T}$ can absorb tokens from all of its precedents, and the number of tokens absorbed from place $p\in P$ is equal to the weight of the edge connecting $p$ and $t$. The number of tokens in each place must be non-negative at any time, hence, a firing of transition $t$ is enabled if and only if the number of tokens in all its precedents $p$ is not smaller than the weight on edge $(p,t)$. Each transition $t\in{T}$ can release tokens to all of its successors, and the number of tokens released to place $p\in P$ is equal to the weight of the edge connecting $t$ and $p$.  It is not necessary that the set of precedents and the set of successors are mutually exclusive. In a TPN, the duration between the moment of absorption and the moment of release is not always negligible, and the duration is called \textit{firing time}. This work considers firing time non-negative and that can follow arbitrary stochastic distribution. The distribution of tokens in the places is called \textit{markings}. The initial state of the system, i.e., the state at time zero, is represented by \textit{the initial marking}. Formally, a TPN can be defined as a 6-tuple $N=(P, T, F,W, M_0, \tau)$, where
\begin{tabular}{ll}
	${P}$ & denotes the set of places;\\
	${T}$ & denotes the set of transitions;\\
	${F}$ & $\subseteq(P\times T)\cup(T\times P)$ denotes the set of edges;\\
	$W:$ & $F \rightarrow \mathbb{Z}^{+}$ denotes the weights on edges;\\
	$M_0:$ & $P\rightarrow \mathbb{N}$ denotes the initial marking;\\
	$\tau:$ & $T\rightarrow random\ distribution$ denotes the firing time of transitions.\\
\end{tabular}

An illustrative example using G/G/2 queue is showed in Figure \ref{fig:gg1_petri}. In a G/G/2 queue, customers arrive at the queue following an arbitrary arrival process. One of the two identical servers (or the idle one if only one is idle) processes server processes customers in the queue one by one, and the processing time is arbitrarily distributed. After processing, the customer will be released from the system immediately, and the server becomes idle. On the right side of Figure \ref{fig:gg1_petri}, the graph of the TPN of G/G/2 is presented. A token in place $p_1$ shows that it is possible to schedule the arrival of next customer, and the firing time of transition $t_1$ is equal to the inter-arrival time. At the finish of firing $t_1$, a customer arrives, and it is possible to schedule the next arrival, thus, it releases one token to $p_1$. Furthermore, an arrival will increase the length of the queue by 1, which is represented by the number of tokens in place $p_2$. Two tokens in place $p_3$ shows that two servers are idle. The firing of transition $t_2$ shows the processing of a customer, and it is enabled if and only if at least one server is idle and there is a customer in the queue. Once the process of the customer finishes, i.e., the finish of firing $t_2$, a server is released, and a token is sent to place $p_3$.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/GG1_petri.png}
	\caption{G/G/2 queue and its TPN model.}
	\label{fig:gg1_petri}
\end{figure}

\subsection{Simulating TPNs}\label{sec:ModelSim}
The event-scheduling approach is the logic behind all the major DES software and used by practitioners when developing simulation codes with general purpose languages (\cite{law2014simulation}). For sake of completeness, the logic for simulating TPN is briefly described. Within the event scheduling worldview of discrete event simulation (\cite{zeigler2018theory}), an event is first scheduled and occurs afterwards. 
An \textit{event list} stores all the scheduled events that will occur in the future. The event with the earliest occurring time is polled to occur from the event list, and the clock is advanced to its occurring time. The system state is changed upon the event occurrence, and the new state enables the schedule of new events. A new cycle starts from polling the earliest event. In TPN, events, whose occurrence changes the system state instantaneously, are either start or finish of transition firings in a TPN, since they are the only moments when the marking is changed. Thus, an event in a simulation realization of TPN can be represented by a triple $(t,type,time)$, where $t$ is a transition, $type$ is either \textit{start} or \textit{finish}, and $time$ is the occurring time. Given a TPN $N=(P, T, F,W, M_0, \tau)$, its simulation can be implemented as in Algorithm \ref{chtp6:Algo:SimulatePetriNet}.

\begin{algorithm}
	\caption{Simulating a TPN.}
	\begin{algorithmic}[1]
		\label{chtp6:Algo:SimulatePetriNet}
		\REQUIRE ~~ \\
		Initial markings: $M_0$. \\

@@

		\WHILE{stopping condition is not true}
		\FOR{each transition $t_{1}\in T$}
		\WHILE{current marking enables the firing of $t_1$, i.e., $M(p_{1})\ge W(p_1,t_{1}),\ \forall (p_1,t_{1})\in F$}
		%\STATE \textbf{Start to fire the enabled transition.}
		%\STATE Add event $e_{1}=(t^{'},start,clock,i^{t^{'},start})$ to set $V$, add arc $(\hat{e},e_1)$ to the set $A$; $i^{t^{'},start}\leftarrow i^{t^{'},start}+1$. 
		%\STATE Among the precedent places of $t^{'}$, find the place $p$ such that $(t,p),(p,t^{'})\in F$, if $\bigg( \left \lceil \frac{W(p,t^{'})i-(\bar{x}^p+1)}{W(t,p)} \right\rceil - \left \lceil \frac{W(p,t^{'})i-\bar{x}^p}{W(t,p)} \right\rceil  \bigg)$ is not equal to zero, add arc $(\hat{e},e_1)$ to set $\Upsilon^p$.
		\STATE Sampling the firing time $\tau_0$ of random variable $\tau(t_1)$
		\FOR{each precedent place $p_1$ of $t_1$}
		\STATE Remove $W(p_1,t_1)$ tokens from $p_1$: $M(p_1) \leftarrow M(p_1)-W(p_1,t_1)$
		\ENDFOR
		%\STATE \textbf{Schedule the end-of-firing event.}
		\STATE Add event $(t_1,finish,clock+\tau_0)$ to event list.
		%\STATE Add event $e_2$ to set $V$, add arc $(e_1,e_2)$ to the set $A$.
		\ENDWHILE
		\ENDFOR
		%\STATE \textbf{Execute the earliset event.}
		\STATE From \textit{EventList}, take the event $(t_2,finish,time)$ with the earliest time. 
		\STATE Advance the clock: $clock\leftarrow time$.
		%\STATE Update $e$ and index: $e\leftarrow(t,end,time,i^{t,end})$, $i^{t,end}\leftarrow i^{t,end}+1$.
		%\STATE Update last occurring event $\hat{e}\leftarrow e$.
		\FOR{each successive place $p_2$ of $t_2$}
		\STATE Add $W(t_2,p_2)$ tokens to $p_2$: $M(p_2) \leftarrow M(p_2)+W(t_2,p_2)$.
		\ENDFOR		
		\STATE Update iteration counter: $k = k+1$
		\ENDWHILE		
	\end{algorithmic}
\end{algorithm}

In Algorithm \ref{chtp6:Algo:SimulatePetriNet}, event list, clock time and system state (i.e., the markings) are initialized as empty, zero and initial markings, respectively. Each iteration is composed of two steps, i.e., to start all the enabled transitions and to finish the firing in the event list with the earliest finishing time. In the first step, if firing of a certain transition $t_1$ is enabled by the current markings, the firing can be started and tokens are immediately removed from precedent places of $t$. In the mean time, the finish of the firing can be scheduled with a delay equal to the firing time, which is a sample $\tau_0$ of random distribution $\tau(t_1)$, and the finish of firing is added to the event list. It can be noticed that the event list contains only finish-of-firing events, since all the start-of-firing events are executed immediately without going into the list. In the second step, the firing with the earliest finish time will be taken from the event list, and the tokens are released to the successive places. 
The iterative procedure stops when a given condition is reached, which is usually a limit of iteration numbers or clock time. %Algorithm \ref{chtp6:Algo:SimulatePetriNet} considers a nondeterministic policy, i.e., if multiple transitions are enabled at the same time, they will fire in any order. One can also define the priority among transitions. In this case, in line 4 of Algorithm \ref{chtp6:Algo:SimulatePetriNet}, transitions with higher priority will be visited first.

\section{MPR models} \label{sec:TPN_MPR}

This section introduces the MPR translating the dynamics of the TPN, equivalent to Algorithm 1. Specifically, the time when transition firings start and finish and markings can all be seen in the MPR as decision variables. The time when transition $t$ starts to fire for the $i$-th time is denoted by $e^{t,s}_i$, and $e^{t,f}_i$ denotes the time when the fire finishes. Variables $\mathcal{E}_k$ denotes the simulation clock at the beginning of iteration $k$ in Algorithm 1. Variables $e^{t,s}_{i}$, $e^{t,f}_{i}$ and $\mathcal{E}_k$ are all real-valued and non-negative. Variable $m^p_k$ is used to denote the number of tokens in place $p$ at the beginning of iteration $k$ in Algorithm 1. Some binary variables are also used in the MPR, and they will be introduced in the following, during the explanation of the model.

>>>>

\title{\Large Generation of Mathematical Programming Representation for Discrete Event Simulation Models of Timed Petri Nets}

@@

\bigskip
\begin{abstract}
This work proposes a mathematical programming (MP) representation of discrete event simulation of timed Petri nets (TPN). Currently, mathematical programming techniques are not widely applied to optimize discrete event systems due to the difficulty of formulating models capable to %catch 
correctly represent the system dynamics. This work connects the two fruitful research fields, mathematical programming and timed Petri nets. In the MP formalism, the decision variables of the model %are composed of 
correspond to the transition firing times and the markings of the TPN, whereas the constraints represent the state transition logic and temporal sequences among events. The MP model and a simulation run of the TPN are then totally equivalent, which is validated %with several cases. 
through several applications. Using a TPN model as input, the MP model can be routinely generated and used as white box for further analysis such as sensitivity analysis, cut generation in optimization procedures, and proof of formal properties.

@@

Discrete event dynamic systems (DEVS) have wide application in manufacturing and service applications. Even though many theoretical and analytical studies on discrete event dynamic systems have been developed, performance of real systems with high degree of complexity can only be evaluated with discrete event simulation (DES). Hence, simulation--optimization algorithms are widely used when numerical performance evaluation must be coupled with optimization, i.e., when the best system configuration, according to some criteria, has to be found meanwhile guaranteeing a given value of some performance measures \citep{fu20115book}.

Most of the simulation--optimization studies consider DES as a \textit{black--box} function. Under the black--box setting, simulation and optimization modules are %decoupled, except that simulation module needs to provide the performance measures of certain solutions when the optimization module queries. With quite limited information from simulation, the optimization module must have many trials, so the black--box procedure can be computationally inefficient.
decoupled. The simulation module provides the performance measure of a given solution generated by the optimization module, which can help a further optimization round. However, the information given by simulation in a black--box setting is quite limited. Hence, the optimization module is almost blind and cannot be very effective, which leads to possibly many trials. This is the main cause for large computational inefficiencies of the overall black--box procedure.

A promising direction to improve the %inefficiency 
efficiency of black--box procedure is to merge the system dynamics into optimization, i.e., considering DES as \textit{white--box} and modeling system dynamics as part of mathematical programming (MP) models. Several works studying such issue can be found in the literature. To cite a few examples, an MP model was proposed to minimize the makespan of single--server manufacturing systems in \citep{di2020makespan}, the buffer allocation problem in multi--stage production flow lines with blocking is addressed in \citep{matta2008simulation}, \citep{weiss2015buffer} and \citep{alfieri2020time}, the optimization of multi stage systems with complex blocking mechanisms are studied in \citep{pedrielli2015integrated}. Decomposition or linear approximation approaches are applied to solve the MP models, %and it can be rather efficient than black-box algorithms and the optimality is also enhanced 
which improve the efficiency of black--box algorithms, thus enhancing also optimality \citep{weiss2015buffer}\citep{alfieri2020time}.

However, white--box simulation--optimization approaches are not as widely--applied as black--box ones. The reason is that developing a white--box algorithm requires %one 
to master discrete event dynamic systems, %its
simulation and mathematical programming, %but those
and these disciplines are not tightly connected. The above mentioned works are all tailored to specific cases, and there is a lack of generalization to extend %the works 
them to other problems. To fill this gap, this work proposes an approach to translate the simulation of a timed Petri net (TPN) into MP models.

In the literature, a few works address similar problems, i.e.,  translating the behavior of general discrete event dynamic systems into mathematical programming representation. \cite{bemporad1999control} proposed a mixed integer programming modeling framework for hybrid systems, whose states are mixed integer, in discrete time. Differently, this work deals with continuous--time discrete--state systems. \cite{chan2008optimization} proposed a modeling framework to translate a DES model into an MP model starting from an Event Relationship Graph (ERG) representation of the system dynamics. However, despite its generality the ERG representation is not as commonly--used as TPNs. Indeed, developing an ERG model is time consuming and %and the 
as a consequence their approach is not used often. On the other hand, TPNs are well--known representations taught in most engineering study courses, and software tools for easy creation of TPN models also exist.

The benefits of using MP models to represent discrete event systems are many. As already mentioned, %above,
when coupled with optimization, the MP-based algorithms can be faster and %gain 
reach better solution than black--box optimization algorithms. Furthermore, black--box approaches have limited capability in solving constrained optimization problems, %but 
while MP--based approaches can easily deal with them \citep{zhang2020models}. The vast theoretical and methodological results developed in the MP field can be introduced into the study of DEVS through simulation. For instance, using sensitivity analysis of MP models, gradient estimates can be easily derived as suggested in \citep{chan2008optimization}. In fact, %we do not proposed 
this paper does not propose to totally replace simulation with MP, but to enrich the toolbox for analyzing and optimizing DEVS. Finally, the application of MP models of DEVS is not limited to optimization but also to show formal properties of the system under study. For instance, \cite{basile2012k} proposed the sufficient and necessary condition of K-diagnosability of TPN based on MP analysis.

The major concern about the application of MP is the computational complexity, %and the solving procedure
as the solution procedure can be time--consuming or even unbearable. However, many approaches from optimization community are available to improve the efficiency. Linear programming approximation \citep{alfieri2012mathematical}, Benders decomposition \citep{weiss2015buffer}, row--column generation \citep{alfieri2020time}, have been studied to solve optimization problems in %reality
real contexts based on mathematical programming representation of DEVS.

The contribution of this work is to propose the first framework translating TPNs into MP models. The translation procedure can be conveniently implemented in general--purpose programming languages and turned into an automated procedure. This work lays a foundation %of 
for developing MP--based approaches %for analyzing and optimizing 
that can be used to analyze and optimize TPNs, based on the vast literature available in the mathematical programming field. 
%The proposed MP models can also be easily transformed into an MPR for optimizing the design parameters of the DES, which are common optimization problems in operations management field.

The rest of the paper is organized as follows. Section \ref{sec:sim} briefly introduces the TPN and its simulation algorithm. Section \ref{sec:TPN_MPR} describes the generation of the MP model from an existing TPN model. Section XXX shows several examples of application and validation of the equivalence between simulation and the MP. Discussion and conclusion are reported in Section XXX and XXX, respectively.

@@

\section{Timed Petri Nets} \label{sec:sim}

\subsection{Basic definitions}
A TPN is a directed bipartite graph. The set of nodes in that graph is partitioned into a set of \textit{places} $P$ and a set of \textit{transitions} $T$. Places and transitions are connected with weighted and directed edges. In the example shown in the right side of Figure \ref{fig:gg1_petri}, places (i.e., $p_1$, $p_2$, $p_3$), transitions (i.e., $t_1$, $t_2$) and tokens are represented by white circles, rectangles and small black circles, respectively. Places hold \textit{tokens}, and tokens are transferred through transition \textit{firings}, i.e., one transition firing absorbs tokens from some places and releases them to some other places. Each transition $t\in{T}$ can absorb tokens from all of its precedents, and the number of tokens absorbed from place $p\in P$ is equal to the weight of the edge connecting $p$ and $t$. The number of tokens in each place must be non--negative at any time; hence, a firing of transition $t$ is enabled if and only if the number of tokens in all its precedents $p$ is not smaller than the weight on edge $(p,t)$. Each transition $t\in{T}$ can release tokens to all of its successors, and the number of tokens released to place $p\in P$ is equal to the weight of the edge connecting $t$ and $p$.  It is not necessary that the set of precedents and the set of successors are mutually exclusive. 
\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/GG1_petri.png}
	\caption{G/G/2 queue and its TPN model.}
	\label{fig:gg1_petri}
\end{figure}
In a TPN, the duration between the moment of absorption and the moment of release is not always negligible, and the duration is called \textit{firing time}. This work considers firing time %non-negative and that
that are non--negative and can follow arbitrary stochastic distribution.

The distribution of tokens in the places is called \textit{markings}. The initial state of the system, i.e., the state at time zero, is represented by \textit{the initial marking}.

Formally, a TPN can be defined as a 6--tuple $N=(P, T, F,W, M_0, \tau)$, where ${P}$ denotes the set of places, ${T}$ denotes the set of transitions, ${F}\subseteq(P\times T)\cup(T\times P)$ denotes the set of edges, $W:F \rightarrow \mathbb{Z}^{+}$ denotes the weights on edges, $M_0:P\rightarrow \mathbb{N}$ denotes the initial marking,
	$\tau:T\rightarrow random\ distribution$ denotes the firing time of transitions.

The illustrative example showed in Figure \ref{fig:gg1_petri} is a G/G/2 queue, in which customers arrive at the queue following an arbitrary arrival process. One of the two identical servers (or the idle one if only one is idle) processes customers in the queue one by one, and the processing time is arbitrarily distributed. After processing, the customer will be immediately released from the system, and the server becomes idle. At the right side of Figure \ref{fig:gg1_petri}, the TPN graph of the G/G/2 system is %presented. 
reported. A token in place $p_1$ shows that it is possible to schedule the arrival of the next customer, and the firing time of transition $t_1$ is equal to the inter-arrival time. At the finish of firing $t_1$, a customer arrives, and it is possible to schedule the next arrival, thus, it releases one token to $p_1$. Furthermore, an arrival will increase the length of the queue by 1, which is represented by the number of tokens in place $p_2$. Two tokens in place $p_3$ shows that two servers are idle. The firing of transition $t_2$ shows the processing of a customer, and it is enabled if and only if at least one server is idle and there is a customer in the queue. Once the process of the customer finishes, i.e., the finish of firing $t_2$, a server is released, and a token is sent to place $p_3$.

\subsection{Simulation of Timed Petri Nets}\label{sec:ModelSim}
The event-scheduling approach is the logic behind all the major DES software and used by practitioners when developing simulation codes with general purpose languages \citep{law2014simulation}. For sake of completeness, the logic for simulating TPN is briefly described in this section. This description will also be useful in the next section to better point out the consistency of the generated MP models with the system dynamics.

Within the event scheduling worldview of discrete event simulation \citep{zeigler2018theory}, an event is first scheduled and occurs afterwards. 
An \textit{event list} stores all the scheduled events that will occur in the future. The event with the earliest occurring time is polled to occur from the event list, and the clock is advanced to its occurring time. The system state is changed upon the event occurrence, and the new state enables the schedule of new events. A new cycle starts from polling the earliest event. In TPN, events, whose occurrence changes the system state instantaneously, are either start or finish of transition firings in a TPN, since they are the only moments when the marking is changed. Thus, an event in a simulation realization of TPN can be represented by a triple $(t,type,time)$, where $t$ is a transition, $type$ is either \textit{start} or \textit{finish}, and $time$ is the occurring time.

Given a TPN $N=(P, T, F,W, M_0, \tau)$, its simulation can be implemented as in Algorithm \ref{chtp6:Algo:SimulatePetriNet}. 
\begin{algorithm}
	\caption{Simulating a TPN.}
	\begin{algorithmic}[1]
		\label{chtp6:Algo:SimulatePetriNet}
		\REQUIRE ~~ \\
		Initial markings: $M_0$. \\

@@

		\WHILE{stopping condition is not true}
		\FOR{each transition $t_{1}\in T$}
		\WHILE{current marking enables the firing of $t_1$, i.e., $M(p_{1})\ge W(p_1,t_{1}),\ \forall (p_1,t_{1})\in F$}
		%\STATE \textbf{Start to fire the enabled transition.}
		%\STATE Add event $e_{1}=(t^{'},start,clock,i^{t^{'},start})$ to set $V$, add arc $(\hat{e},e_1)$ to the set $A$; $i^{t^{'},start}\leftarrow i^{t^{'},start}+1$. 
		%\STATE Among the precedent places of $t^{'}$, find the place $p$ such that $(t,p),(p,t^{'})\in F$, if $\bigg( \left \lceil \frac{W(p,t^{'})i-(\bar{x}^p+1)}{W(t,p)} \right\rceil - \left \lceil \frac{W(p,t^{'})i-\bar{x}^p}{W(t,p)} \right\rceil  \bigg)$ is not equal to zero, add arc $(\hat{e},e_1)$ to set $\Upsilon^p$.
		\STATE Sample the firing time $\tau_0$ of random variable $\tau(t_1)$
		\FOR{each precedent place $p_1$ of $t_1$}
		\STATE Remove $W(p_1,t_1)$ tokens from $p_1$: $M(p_1) \leftarrow M(p_1)-W(p_1,t_1)$
		\ENDFOR
		%\STATE \textbf{Schedule the end-of-firing event.}
		\STATE Add event $(t_1,finish,clock+\tau_0)$ to event list.
		%\STATE Add event $e_2$ to set $V$, add arc $(e_1,e_2)$ to the set $A$.
		\ENDWHILE
		\ENDFOR
		%\STATE \textbf{Execute the earliset event.}
		\STATE From \textit{EventList}, take the event $(t_2,finish,time)$ with the earliest time. 
		\STATE Advance the clock: $clock\leftarrow time$.
		%\STATE Update $e$ and index: $e\leftarrow(t,end,time,i^{t,end})$, $i^{t,end}\leftarrow i^{t,end}+1$.
		%\STATE Update last occurring event $\hat{e}\leftarrow e$.
		\FOR{each successive place $p_2$ of $t_2$}
		\STATE Add $W(t_2,p_2)$ tokens to $p_2$: $M(p_2) \leftarrow M(p_2)+W(t_2,p_2)$.
		\ENDFOR		
		\STATE Update iteration counter: $k = k+1$
		\ENDWHILE		
	\end{algorithmic}
\end{algorithm}
In Algorithm \ref{chtp6:Algo:SimulatePetriNet}, event list, clock time and system state (i.e., the markings) are initialized as empty, zero and initial markings, respectively. Each iteration is composed of two steps, i.e., to start all the enabled transitions and to finish the firing in the event list with the earliest finishing time. In the first step, if firing of a certain transition $t_1$ is enabled by the current markings, the firing can be started and tokens are immediately removed from precedent places of $t$. In the mean time, the finish of the firing can be scheduled with a delay equal to the firing time, which is a sample $\tau_0$ of random distribution $\tau(t_1)$, and the finish of firing is added to the event list. It can be noticed that the event list contains only finish--of--firing events, since all the start--of--firing events are executed immediately without going into the list. In the second step, the firing with the earliest finish time will be taken from the event list, and the tokens are released to the successive places. 
The iterative procedure stops when a given condition is reached, which is usually %a limit of iteration numbers or clock time. 
a given number of iterations or a value of the clock time.

%Algorithm \ref{chtp6:Algo:SimulatePetriNet} considers a nondeterministic policy, i.e., if multiple transitions are enabled at the same time, they will fire in any order. One can also define the priority among transitions. In this case, in line 4 of Algorithm \ref{chtp6:Algo:SimulatePetriNet}, transitions with higher priority will be visited first.

\section{Generation of Mathematical Programming Representations} \label{sec:TPN_MPR}
